Depletion capabilities in the OpenMC Monte Carlo particle transport code

### 2020-07-01 manuscript.tex
Address @drewejohnson and @liangjg feedback

Monte Carlo (MC) particle transport simulations are now widely used in reactor
design and analysis in a variety of roles thanks to continued advances in
computing. The earliest efforts to couple MC particle transport with depletion
date back to the 1990s~\citep{moore1995inel,trellue1998lanl}. These early
efforts involved taking existing MC and depletion codes, such as
MCNP~\citep{goorley2012nt} and ORIGEN~\citep{croff1983nt}, and putting a
``wrapper'' around them that orchestrated the transfer of solutions from one
code to the other. Thus, the actual transport/depletion codes themselves
required no source code changes. In the last decade, many MC code developers
have chosen to directly include a depletion solver as a first-rate feature,
including MCNPX~\citep{waters2007aip} (which has since been merged into
MCNP6~\citep{goorley2012nt}), Serpent~\citep{leppanen2015ane}, and
MC21~\citep{griesheimer2015ane}.

@@

The equation that governs the transmutation of nuclides by nuclear reactions and
radioactive decay inside of an irradiated environment can be written as
\begin{equation}
  \label{eq:depletion}
  \begin{split}
  \frac{dN_i(t)}{dt} = &\sum\limits_j \left [ f_{j \rightarrow i}
  \int_0^\infty dE \; \sigma_j (E, t) \phi(E,t) + \lambda_{j\rightarrow i}
  \right ] N_j(t) \\ &- \left [\int_0^\infty dE \; \sigma_i (E,t) \phi(E,t) +
  \sum\limits_j \lambda_{i\rightarrow j} \right ] N_i(t), \quad i=1,\dots,n
  \end{split}
\end{equation}
where
\begin{equation*}
  \begin{split}
  N_i(t) &\equiv \text{density of nuclide $i$ at time $t$} \\
  \sigma_i(E,t) &\equiv \text{transmutation cross section for nuclide $i$ at energy $E$ and time $t$} \\
  \phi(E,t) &\equiv \text{neutron flux at energy $E$ and time $t$} \\
  f_{j \rightarrow i} &\equiv \text{fraction of transmutation reactions in nuclide $j$ that produce nuclide $i$} \\
  \lambda_{j \rightarrow i} &\equiv \text{decay constant for decay modes in nuclide $j$ that produce nuclide $i$} \\
  n &\equiv \text{total number of nuclides.}
  \end{split}
\end{equation*}
Note that we have not included the spatial dependence of the flux or cross
sections. For \cref{eq:depletion} to be accurate, the spatial variation of the
reaction rates over the region which is being depleted should be small. For a
reactor model, this may entail using a unique material in each fuel pin that is
depleted separately. Ultimately, the user is responsible for deciding what level
of spatial discretization is adequate for the problem at hand.

@@

Using all incident neutron, decay, and FPY sublibrary files from an ENDF-format
library, such as ENDF/B-VIII.0, results in a \emph{full} depletion chain that
typically includes thousands of nuclides. Because the use of a chain with $n$
nuclides will result in an $n\times n$ burnup matrix, there is an incentive to
reduce the number of nuclides in the depletion chain to minimize both the size
of the burnup matrix and the number of nuclides that are tracked during a
transport solve. The depletion chain class in OpenMC's Python API includes a
method for automatically reducing the size of the chain. Starting with a set of
nuclides that must be included, the method will recursively go through all decay
and reaction modes to determine what products may be produced. Using this method
on the ENDF/B-VII.1 library starting from a list of all stable or very long
lived (half-life greater than $10^{15}$ sec) nuclides reduces the total number
of nuclides in the depletion chain from 3,820 to 1,693.

@@

The reaction rates calculated by OpenMC are given in units of reactions per
source particle. For depletion, it is necessary to compute an absolute reaction
rate in reactions per second. To do so, the total observed heating in the system
is determined, and then the ratio of the system power to the observed heating is
used as a normalization factor on the reaction rates. OpenMC allows the total
heating to be determined by two different methods. In the first method, fission
reaction rates are multiplied by constant Q values in the depletion chain file
to estimate the total heating. This method is particularly useful for making
inter-code comparisons, where one must ensure that the same Q values for fission
are applied. In the second method, the actual heating rate is tallied over the
entire system (separate from the tally for transmutation reaction rates). This
method makes it possible to account for the energy dependence of the fission
energy release data as well as the fact that neutrons and photons may leak out
of the system, carrying away their energy with them. The latter effect may be
particularly important for small, leaky systems.

>>>>

Monte Carlo (MC) particle transport simulations are now widely used in reactor
design and analysis in a variety of roles thanks to continued advances in
computing. The earliest efforts to couple MC particle transport with depletion
date back to the 1990s~\citep{moore1995inel,trellue1998lanl}. These early
efforts involved taking existing MC and depletion codes, such as
MCNP~\citep{goorley2012nt} and ORIGEN~\citep{croff1983nt}, and putting a
``wrapper'' around them that orchestrated the transfer of solutions from one
code to the other. Thus, the actual transport/depletion codes themselves
required no source code changes. In the last decade, many MC code developers
have chosen to directly include a depletion solver as a first-rate feature,
including MCNPX~\citep{waters2007aip} (which has since been merged into
MCNP6~\citep{goorley2012nt}), Serpent~\citep{leppanen2015ane},
MC21~\citep{griesheimer2015ane}, and RMC~\citep{wang2015ane}.

@@

The equation that governs the transmutation of nuclides by nuclear reactions and
radioactive decay inside of an irradiated environment can be written as
\begin{equation}
  \label{eq:depletion}
  \begin{split}
  \frac{dN_i(t)}{dt} = &\sum\limits_j \left [ f_{j \rightarrow i}
  \int_0^\infty dE \; \sigma_j (E, t) \phi(E,t) + \lambda_{j\rightarrow i}
  \right ] N_j(t) \\ &- \left [\int_0^\infty dE \; \sigma_i (E,t) \phi(E,t) +
  \sum\limits_j \lambda_{i\rightarrow j} \right ] N_i(t), \quad i=1,\dots,n
  \end{split}
\end{equation}
where
\begin{equation*}
  \begin{split}
  N_i(t) &\equiv \text{density of nuclide $i$ at time $t$} \\
  \sigma_i(E,t) &\equiv \text{transmutation cross section for nuclide $i$ at energy $E$ and time $t$} \\
  \phi(E,t) &\equiv \text{neutron flux at energy $E$ and time $t$} \\
  f_{j \rightarrow i} &\equiv \text{fraction of transmutation reactions in nuclide $j$ that produce nuclide $i$} \\
  \lambda_{j \rightarrow i} &\equiv \text{decay constant for decay modes in nuclide $j$ that produce nuclide $i$} \\
  n &\equiv \text{total number of nuclides.}
  \end{split}
\end{equation*}
Note that we have not included the spatial dependence of the flux or cross
sections. For \cref{eq:depletion} to be accurate, the spatial variation of the
reaction rates over the region which is being depleted should be small. For a
reactor model, this may entail using a unique material in each fuel pin that is
depleted separately.\footnote{OpenMC includes functions for automatically
differentiating materials in cells that are repeated throughout a geometry.}
Ultimately, the user is responsible for deciding what level of spatial
discretization is adequate for the problem at hand.

@@

Using all incident neutron, decay, and FPY sublibrary files from an ENDF-format
library, such as ENDF/B-VIII.0, results in a \emph{full} depletion chain that
typically includes thousands of nuclides. Because the use of a chain with $n$
nuclides will result in an $n\times n$ burnup matrix, there is an incentive to
reduce the number of nuclides in the depletion chain to minimize both the size
of the burnup matrix and the number of nuclides that are tracked during a
transport solve. The depletion chain class in OpenMC's Python API includes a
method for automatically reducing the size of the chain. Starting with a set of
nuclides that must be included, the method will recursively go through all decay
and reaction modes to determine what products may be produced. This can help
minimize both the size of the depletion matrix and the number of nuclides for
which reaction rates must be tallied, thereby reducing the time spent in the
depletion and transport solvers. Using this method on the ENDF/B-VII.1 library
starting from a list of all stable or very long lived (half-life greater than
$10^{15}$ sec) nuclides reduces the total number of nuclides in the depletion
chain from 3,820 to 1,693.

@@

The reaction rates calculated by OpenMC are given in units of reactions per
source particle. For depletion, it is necessary to compute an absolute reaction
rate in reactions per second. To do so, the total observed heating in the system
is determined, and then the ratio of the system power to the observed heating is
used as a normalization factor on the reaction rates. OpenMC allows the total
heating to be determined by two different methods. In the first method, fission
reaction rates are multiplied by constant Q values in the depletion chain file
to estimate the total heating. This method is particularly useful for making
inter-code comparisons, where one must ensure that the same Q values for fission
are applied. In the second method, the actual heating rate is tallied over the
entire system (separate from the tally for transmutation reaction rates). This
method makes it possible to account for the energy dependence of the fission
energy release data as well as the fact that neutrons and photons may leak out
of the system, carrying away their energy with them. The latter effect may be
particularly important for small, high-leakage systems.
